Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-04-30 joint_inf.tex
clarify, be explicit, address (most of) reviewer 2

Assume the binary symmetric model, namely with a character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution \citep{Semple2003-em}.
Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
We observe $\nCols$ independent and identically distributed samples of character data, i.e., an alignment with $\nCols$ columns, $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
The corresponding unobserved ancestral states are $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$ and distributed as $\ancestralStateColumnRV$.

@@

Fidelities have useful algebraic properties, and generating probabilities using the Hadamard transform have an especially simple form (see \eqref{eq:hadamard_probability} in the Appendix).
For a four-taxon tree, define the general branch fidelity parameter $t=\{\theta_1,\theta_2,\theta_3,\theta_4,\theta_5\}$ where fidelities are ordered in the order of the taxa with the internal branch last (Fig.~\ref{fig:farris-fels-top}).

@@

The alternative approach \citep{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
This is known in statistics as a profile likelihood \citep{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\tau, t; \fullAlignment) = \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i\in\alphabet^{\nAncestralStateRows}} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \max_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates).
\end{equation}
We use $\hat{\fullAncestralStates}$ to denote an estimate for $\fullAncestralStates$ obtained by maximizing \eqref{eq:profile_likelihood}, and estimate a topology and branch fidelities using this profile likelihood as
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment).
\end{equation}
In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \citep{Sagulenko2017-jo}.

@@

To show an inconsistency in topology estimation, we start with true generating parameters $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ on the Farris topology (Fig.~\ref{fig:farris-fels-top}a).
We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology than for the true, generating Farris topology.
To do so, we construct an upper bound $C_0(x^*, y^*)$ for the likelihood given the Farris topology as a function of $x^*$ and $y^*$ and, similarly, a lower bound $C_1(x^*, y^*)$ for the likelihood given the Felsenstein topology.
When $C_0(x^*, y^*) < C_1(x^*, y^*)$, the likelihood in the Felsenstein case is larger than the likelihood in the Farris case, demonstrating inconsistency (Fig.~\ref{fig:inconsistency-farris}).
\begin{restatable}{theorem}{topoInconsist}
Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x, y, x, y, w\}$.
There exist $C_0(x^*, y^*), C_1(x^*, y^*),$ and a set of $0 < x^*, y^* < 1$ such that
$$
\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t) \le C_0(x^*, y^*),
$$
$$
C_1(x^*, y^*) \le \max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t)
$$
with $C_0(x^*, y^*) < C_1(x^*, y^*)$.
\end{restatable}
The proof of this theorem is by a detailed examination of inequalities.
Intuitively, $\tau_2$ is favored in performing joint inference since the objective function for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its objective function while $\tau_2$ has many more.
This enables more possible maxima for $\tau_2$ even if data are generated from $\tau_1$.

@@

We now consider the problem of branch length estimation on the correct tree using joint estimation.
As described above, we use the equivalent but different notion of branch fidelities.
We analyze two settings on the Farris tree, corresponding to whether some branch fidelities are fixed at their true values or not.
As above, assume that data is generated from the Farris tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).
In the first ``restricted'' case, we show that for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one (zero branch length) instead of its true value of $y^*$.
That is, if we estimate $x$ and $y$ correctly, then, for
$$
\hat{w} = \arg\max_{w} \ \ell_{\tau_1,t^*}(\tau_1, \{x^*,y^*,x^*,y^*,w\}),
$$
there is a set of values for $x^*$ and $y^*$ where $\hat{w}\equiv 1$.
In the general case, we do the same but with
$$
(\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,t^*}(\tau_1, \{x,y,x,y,w\})
$$
to find a region where the inferred values do not converge to the generating values.
These situations are in contrast to the approach using marginal likelihood where $\hat{w}$ necessarily converges to $y^*$ as the number of observations grows \citep{RoyChoudhury2015-ta}.

@@

Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and estimate the internal branch parameter $w$.
\begin{restatable}{theorem}{restrictedBranchInconsist}
\label{thm:restricted-bl}
Let
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
$$
The maximum likelihood value $\hat{w}$ is equal to 1 if
$$
%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
-\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0
$$
and there exists a set of $0 < x^*, y^* < 1$ satisfying this.
\end{restatable}
This theorem allows us to demarcate a region of biased internal branch length estimation by plotting where the inequality in Theorem~\ref{thm:restricted-bl} is satisfied (Fig.~\ref{fig:bl-inconsistency}).
Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch with this branch length estimated to be zero (i.e., has fidelity $\hat{w} = 1$).
As an intuition for the theoretical development, seeing no change along a branch more likely increases the likelihood by introducing a term of $(1+\theta)$ instead of $(1-\theta)$, and branch fidelities will be positively biased due to this.
If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology.

@@

Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same procedure.
For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look reasonable, with $\hat{w}$ increasing as $y^*$ increases, though Fig.~\ref{fig:bl-general-bias} shows there is a systematic positive bias in this procedure even when $\hat{w}$ is not estimated to be one.
To understand the quality of each fit, we report the range of $\hat{w}-y^*$ where $\hat{w}\neq 1$.
For joint inference, the errors range from $[-7\times 10^{-3}, 8\times 10^{-2}]$ and for marginal inference, $[-8\times 10^{-8}, 5\times 10^{-7}]$ showing that, even in cases where joint inference does not estimate $\hat{w}$ to be exactly one, it still fails to achieve a low error from truth.

>>>>

Assume the binary symmetric model, namely with a character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution \citep{Semple2003-em}.
Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
We observe $\nCols$ independent and identically distributed samples of character data, i.e., an alignment with $\nCols$ columns, $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
The corresponding unobserved ancestral states are $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$ and distributed as $\ancestralStateColumnRV$ with each $\ancestralStateColumn_i\in\alphabet^\nAncestralStateRows$.

@@

Fidelities have useful algebraic properties.
As data becomes plentiful, we use the Hadamard transform (see \eqref{eq:hadamard_probability} in the Appendix) to compute the exact probabilities that generate each particular configuration of taxa---we call these ``generating probabilities''---and these have an especially simple form.
For a four-taxon tree, define the general branch fidelity parameter $t=\{\theta_1,\theta_2,\theta_3,\theta_4,\theta_5\}$ where fidelities are ordered in the order of the taxa with the internal branch last (Fig.~\ref{fig:farris-fels-top}).

@@

The alternative approach \citep{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
This is known in statistics as a profile likelihood \citep{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\tau, t; \fullAlignment) = \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i\in\alphabet^{\nAncestralStateRows}} \ \Pr(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) = \max_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates).
\end{equation}
We use $\hat{\fullAncestralStates}$ to denote an estimate for $\fullAncestralStates$ obtained by maximizing \eqref{eq:profile_likelihood}, and estimate a topology and branch fidelities using this profile likelihood as
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment).
\end{equation}
In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities arising from taking maxima row-wise in Tables~\ref{tab:farris_likelihoods} and~\ref{tab:fels_likelihoods} to obtain the likelihood term for each most likely ancestral state, these terms depending on the unknown $(\tau, t)$.
For this reason, in practice, the joint inference strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \citep{Sagulenko2017-jo}.

@@

To show an inconsistency in topology estimation, we start with true generating parameters $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ on the Farris topology (Fig.~\ref{fig:farris-fels-top}a).
We show that, as $\nCols\rightarrow\infty$, there exist values for $x^*$ and $y^*$ such that the value of the likelihood after maximizing using joint inference is greater for the Felsenstein topology ($\tau_2$) than for the true, generating Farris topology ($\tau_1$).
To do so, we construct an upper bound $C_0(x^*, y^*)$ for the likelihood given the Farris topology as a function of $x^*$ and $y^*$ and, similarly, a lower bound $C_1(x^*, y^*)$ for the likelihood given the Felsenstein topology.
When $C_0(x^*, y^*) < C_1(x^*, y^*)$, the likelihood in the Felsenstein case is larger than the likelihood in the Farris case, demonstrating inconsistency (Fig.~\ref{fig:inconsistency-farris}).
\begin{restatable}{theorem}{topoInconsist}
Let $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ and $t=\{x, y, x, y, w\}$.
There exist $C_0(x^*, y^*), C_1(x^*, y^*),$ and a set of $0 < x^*, y^* < 1$ such that
$$
\max_{t} \ \ell_{\tau_1,t^*}(\tau_1, t) \le C_0(x^*, y^*),
$$
$$
C_1(x^*, y^*) \le \max_{t} \ \ell_{\tau_1,t^*}(\tau_2, t)
$$
with $C_0(x^*, y^*) < C_1(x^*, y^*)$.
\end{restatable}
The proof of this theorem is by a detailed examination of inequalities.
Intuitively, $\tau_2$ is favored in performing joint inference since the objective function for $\tau_2$ has more ``degrees of freedom''---Table~\ref{tab:likelihoods} shows that $\tau_1$ has only three possible forms for its objective function while $\tau_2$ has many more.
This enables more possible maxima for $\tau_2$ even if data are generated from $\tau_1$.

@@

We now consider the problem of branch length estimation on the correct tree using joint estimation.
As described above, we use the notion of fidelities to analyze what happens as changes occur along each branch as opposed to the standard branch length formulation.
We analyze two settings on the Farris tree, corresponding to whether some branch fidelities are fixed at their true values or not.
As above, assume that data is generated from the Farris tree with two top branches of fidelity $x^*$ and all other branches of fidelity $y^*$ (Fig.~\ref{fig:farris-fels-top}).
In the first ``restricted'' case, we show that for a nontrivial subset of possible values for $x^*$ and $y^*$, the interior branch fidelity parameter $w$ will be consistently overestimated as exactly equal to one (zero branch length) instead of its true value of $y^*$.
That is, if we estimate $x$ and $y$ correctly, then, for
$$
\hat{w} = \arg\max_{w} \ \ell_{\tau_1,t^*}(\tau_1, \{x^*,y^*,x^*,y^*,w\}),
$$
there is a set of values for $x^*$ and $y^*$ where $\hat{w}\equiv 1$.
In the general case, we do the same but with
$$
(\hat{x}, \hat{y}, \hat{w}) = \arg\max_{x,y,w} \ \ell_{\tau_1,t^*}(\tau_1, \{x,y,x,y,w\})
$$
to find a region where the inferred values do not converge to the generating values.
These situations are in contrast to the approach using marginal likelihood where $\hat{w}$ necessarily converges to $y^*$ as the number of observations grows \citep{RoyChoudhury2015-ta}.

@@

Fix estimated fidelities $x=x^*$ and $y=y^*$ to their true, generating values and estimate the internal branch parameter $w$.
\begin{restatable}{theorem}{restrictedBranchInconsist}
\label{thm:restricted-bl}
Let
$$
\beta := \beta(x^*, y^*) = 1+(x^*)^2+(y^*)^2+(x^*)^2(y^*)^2,
$$
$$
\gamma := \gamma(x^*, y^*) = 4x^*y^*.
$$
The maximum likelihood value $\hat{w}$ is equal to 1 if
$$
%\beta^2-\gamma^2-2\gamma^2\alpha_1-2\gamma^2\alpha_2+2\gamma\alpha_1\beta-2\gamma\alpha_2\beta \ge 0.
-\gamma^2\left(1 + \frac{1}{2}\beta\right) + 2\gamma\beta x^*(y^*)^2 + \beta^2 \ge 0
$$
and there exists a set of $0 < x^*, y^* < 1$ satisfying this.
\end{restatable}
This theorem allows us to demarcate a region of biased internal branch length estimation by plotting where the inequality in Theorem~\ref{thm:restricted-bl} is satisfied (Fig.~\ref{fig:bl-inconsistency}).
Intuitively, this happens when estimated ancestral sequences at internal nodes are identical across a branch with this branch length estimated to be zero (i.e., has fidelity $\hat{w} = 1$).
As an intuition for the theoretical development, if there is no change along a branch with fidelity $\theta$, the likelihood will more likely include a term of $(1+\theta)$ instead of $(1-\theta)$ for this event, resulting in positive bias of branch fidelities.
If we allow multifurcating trees in our inference, then we can think of this as another instance of converging to the wrong topology.

@@

As data were generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same procedure.
For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look reasonable, with $\hat{w}$ increasing as $y^*$ increases, though Fig.~\ref{fig:bl-general-bias} shows there is a systematic positive bias in this procedure even when $\hat{w}$ is not estimated to be one.
To understand the quality of each fit, we report the range of $\hat{w}-y^*$ where $\hat{w}\neq 1$.
For joint inference, the errors range from $[-7\times 10^{-3}, 8\times 10^{-2}]$ and for marginal inference, $[-8\times 10^{-8}, 5\times 10^{-7}]$ showing that, even in cases where joint inference does not estimate $\hat{w}$ to be exactly one, it still fails to achieve a low error from truth.
